A general Nullstellensatz for generalized spaces

### 2019-03-17 paper-qcoh.tex
Cite Blass's /Seven trees in one/
to do
Discuss nontriviality of the Nullstellensatz

The resulting formula~$\varphi'$ is a geometric formula over the signature
of~$\TT/U_\TT(A)$.
The verification of~$A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$
$(\varphi \dashv\vdash \varphi')$}$ rests on the observation
\[ A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\bigl((\exists \vec y_k\_
\theta_k) \seq{[]} \bigvee_\ell \{ \top \,|\, \text{$(\exists \vec y_\ell\_ \theta_\ell)$
holds for~$U_\TT$} \}\bigr)$} \]
which in turn can be checked on the covering~$(\{ \vec y_j\_ \alpha_j \}
\xrightarrow{[\theta_j]} \{ \vec x\_ \alpha \})_j$, applying
Lemma~\ref{lemma:truth-to-provability} and using that~$\TT$ (and
hence~$\ul{\TT}$) proves~$(\exists \vec y_j\_ \theta_j) \wedge (\exists \vec
y_k\_ \theta_k) \seq{\vec x} \bigvee \{ \top \,|\, j = k \}$.
\end{proof}

@@

\begin{rem}Theorem~\ref{thm:nullstellensatz} cannot be strengthened to
arbitrary geometric sequents. For instance, in the case that~$\TT$ is the
theory of rings, the internal geometric sequent~$(\top \seq{x\?U_\TT}
\bigvee_{a\?U_\TT} (x = e_a))$ trivially holds for~$U_\TT$. However, it is not
provable modulo~$\TT/U_\TT$, as for instance the polynomial algebra~$U_\TT[X]$
does not validate it.\end{rem}

@@

TODO:

* XXX Nichttrivialität des Nullstellensatzes diskutieren (Spezialfall top |- φ
  betrachten, dieser ist nämlich schon mehr oder weniger trivial)

>>>>

The resulting formula~$\varphi'$ is a geometric formula over the signature
of~$\TT/U_\TT(A)$.
The verification of~$A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$
$(\varphi \dashv\vdash \varphi')$}$ rests on the observation
\[ A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\bigl((\exists \vec y_k\_
\theta_k) \seq{[]} \bigvee_\ell \{ \top \,|\, \text{$(\exists \vec y_\ell\_ \theta_\ell)$
holds for~$U_\TT$} \}\bigr)$} \]
which in turn can be checked on the covering~$(\{ \vec y_j\_ \alpha_j \}
\xrightarrow{[\theta_j]} \{ \vec x\_ \alpha \})_j$, applying
Lemma~\ref{lemma:truth-to-provability} and using that~$\TT$ (and
hence~$\ul{\TT}$) proves~$(\exists \vec y_j\_ \theta_j) \wedge (\exists \vec
y_k\_ \theta_k) \seq{\vec x} \bigvee \{ \top \,|\, j = k \}$. (This kind of
reasoning also appears, for instance, in~\cite[p.~20]{blass:seven-trees}.)
\end{proof}

@@

The force of the Nullstellensatz of Theorem~\ref{thm:nullstellensatz} is for
sequents~$\sigma$ which are not of the form~$(\top \seq{[]} \varphi)$. Indeed,
for such sequents, even~$\TT$ proves~$(\speak{$\sigma$ holds for~$U_\TT$}
\Rightarrow \speak{$\ul{\TT}/U_\TT$ proves~$(\top \seq{[]} \varphi)$})$, as can
be seen by a simple induction on the structure of~$\varphi$. For more general
sequents, Theorem~\ref{thm:nullstellensatz} expresses a nontrivial fact: that
truth of a conditional, universally-quantified statement implies a single
unconditional statement.

\begin{rem}Theorem~\ref{thm:nullstellensatz} cannot be strengthened to
arbitrary geometric sequents. For instance, in the case that~$\TT$ is the
theory of rings, the internal geometric sequent~$(\top \seq{x\?U_\TT}
\bigvee_{a\?U_\TT} (x = e_a))$ trivially holds for~$U_\TT$. However, this sequent is not
provable modulo~$\ul{\TT}/U_\TT$, as for instance the polynomial algebra~$U_\TT[X]$
does not validate it.\end{rem}

@@

* XXX Cite Filip Bár?
